**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-06-10 submission/manuscript.Rmd
table S1

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed (suboptimal) modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

- Without rigorous pipelines, we can't believe the machine learning models.

- To speed up progress, let's slow down and make it right.

- Let's define and start using good ML practices.

@@

For regularized logistic regression and SVM with linear kernel we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero), removing features that could be correlated with other important features. Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.

@@

We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`). The median AUROC of therandom forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].

@@

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models. We examined the 5 predictive (highest weight) and protective (lowest weight) features identified by the model. We calculated the mean weights and coefficients of these features over the 100 data-splits. In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3]. We explained the feature importances in non-linear models using permutation importance on the held-out test data where we permuted groups of correlated features (see methods). For the tree-based models, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B]. Other OTUs had none to minimal effect on the predictive performance.

- decision tree has no rgularization.

@@

- Testing with each model will take the same amount of time but if you are training a model and optimize hyper-parameters, then you need to consider the computational burden. The computational efficiency of different models have been investigated before. Explain for each model. In tree based models, depth of tree, number of features, .. matter. L1 linear is very fast.

- But we can reduce time paralellizing different components of the pipeline. Here I have parallelized each datasplit but we can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.

@@

 - decision tree and l1 why so low?

>>>>

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies. However, the results so far are overoptimistic both in terms of prediction performance and validity of developed ML models. Without rigorous ML pipelines, we cannot trust ML models. Before we can speed up progress, we need to slow down, define and start using good ML practices.

@@

For regularized logistic regression and SVM with linear kernel we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero). Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.

@@

We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].

@@

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609) and genera *Gamella*  (OTU00426) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models. We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC were overl in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B]. Other OTUs had none to minimal effect on the predictive performance.

@@

As the complexity of a ML model increase, its training times will increase (Table S1). When regularization and hyperparameter optimization (i.e. model selection) steps are added to the model pipeline, training times will increase further. The more hyperparameters a model has, the longer it will take to train. The computational burden that comes from model complexity, can be reduced by parallelizing different segments of the ML pipeline. In this study we have parallelized each datasplit which allowed 100 atasplits to be run at the same time for each model. We can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.

@@

 - decision tree and l1 why so low?  Decision tree model has no regularization which is why the AUROC values were low.
